Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
move insert to section templates

\begin{figure}[h]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
		\label{traverse_lock_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		\label{traverse_lock_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms}
	\label{traverse_lock}
\end{figure}

@@

@@

\than{I moved \texttt{insert} from Section 4 to here. William: Could you please write something here to have a smooth transition from \texttt{traverse} to \texttt{insert}, or do you think you need to have subsections for \texttt{traverse} and \texttt{insert}? }

\begin{figure}[h]
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle, escapechar=|]{lock_insert.c} 
		\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
		\label{insert_lock}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
		\caption{The \lstinline{insert} method of the give-up template algorithm}
		\label{insert_giveup}
	\end{subfigure} 
	\caption{The \lstinline{insert} method of the lock-coupling and give-up template algorithms}
	\label{insert_lock_giveup} 
\end{figure}

\than{Should we move the code to template section?}

Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling template. %\wm{Actually, is this \lstinline{insert} function part of the template? If so, should we discuss it in section 3?} 
It makes use of the \texttt{traverse} function %of Section \ref{lock-coupling-algo} 
to find the node at which to insert the key \lstinline{k}. If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the data-structure-specific \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}).

@@

Figure \ref{proof_lock_insert} in Appendix \ref{sec:apd_proof} depicts the \texttt{insert} function within the lock-coupling template for BST, annotated using separation logic.
The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{p}$. We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{proof_lock_insert}). After the lock for the root node is acquired (line 5 in Figure \ref{proof_lock_insert}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{r})$. We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
		\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
		\right\rangle
	}
	\\ 
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
		\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
			\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
		\end{array}
		\ \Bigg| \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
In the $\mathsf{then}$ branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
we know that $\texttt{pn->n'->t} \neq \texttt{NULL}$, i.e., the key to be inserted already exists in the tree. In this scenario, we update the node's value while maintaining the rest of the tree structure (line 9 in Figure \ref{proof_lock_insert}). Otherwise, when \texttt{traverse} reaches a leaf and produces $\mathit{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 14-15 in Figure \ref{proof_lock_insert}). We can then call the \texttt{insertOp} specification, adding the new key-value pair to the tree. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree. 
We can then release the lock (line 21 in Figure \ref{proof_lock_insert}) return the node's contents to the shared state. %Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
%\wm{Do we ever define $\treerep$?}

@@

\begin{figure}[!ht]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in an ML-like language}
		\label{traverse_giveup_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
		\label{traverse_giveup_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the give-up template algorithms}
	\label{traverse_giveup}
\end{figure}

@@

@@

\than{I moved \texttt{insert} from Section 4 to here. William: Could you please write something here to have a smooth transition from \texttt{traverse} to \texttt{insert}, or do you think you need to have subsections for \texttt{traverse} and \texttt{insert}? }

The give-up version of the \texttt{insert} method (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version. The one difference is that it does not require acquiring a lock before calling its \texttt{traverse} function. The proof of the \lstinline{insert} function is similar to that of the previous function: the only difference is that the give-up version of \lstinline{insert} does not acquire a lock before calling \lstinline{traverse}, reflecting the difference in the locking patterns and the preconditions of \lstinline{traverse} between the two templates. In Appendix \ref{sec:apd_proof}, Figure \ref{proof_giveup_insert} presents an annotation that uses separation logic for the \texttt{insert} function within the give-up template for BST.

@@

The give-up BST uses the same definition and specification of \texttt{findNext}, while introducing an extra helper function called \texttt{inRange} \wm{If \lstinline{inRange} is part of the template, I think we should discuss it in the previous section}. The \texttt{insertOp} function is slightly different from that of the lock-coupling template, as each \lstinline{node_t} struct has two additional fields, \texttt{min} and \texttt{max}, to implement \lstinline{inRange}, as described in Section \ref{give-up-algo}.

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% inRange for give-up
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{inRange(pn, k)} 
		\\
		{\color{blue}
			\left\{\mathit{res.} \
			\begin{array}{l}
				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} \  \ast  \ 
				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{min} < \texttt{k} < \texttt{max})\ \mathsf{else}\ (\texttt{k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max}))
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specifications of \texttt{inRange} for the give-up template}
	\label{fig:inRange_giveup}
\end{figure}

The \texttt{inRange} function (Figure \ref{fig:inRange_giveup}) simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.

\subsection{Lock-coupling BST}
\label{BST_lock_insert}
\than{What do we need to say here? (1) - still keep subsection and filled \texttt{insertOp} and the way we prove \texttt{insert} here, or (2) - get rid of this subsection.}
\subsection{Give-up BST}
\label{BST_giveup_insert}
\than{What do we need to say here? (1) - still keep subsection and filled \texttt{insertOp}, and \texttt{inRange} and the way we prove \texttt{insert} here, or (2) - get rid of this subsection.}

@@

Even though the proof for the \texttt{insert} function within the lock-coupling templates was line-of-code (LoC) intensive, future enhancements could significantly streamline this process. Otherwise, the construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, necessitated a considerable amount of time and effort.
However, we expect that functions such as \texttt{traverse}, \texttt{inRange}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.  The methods \texttt{findNext} and \texttt{insertOp} may also be adapted to other data structures, such as linked lists, with some additional effort. 
Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext} and \texttt{insertOp} to restructure the \texttt{node} struct. While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future. The full development of our mechanization effort is available online
\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.

>>>>

\begin{figure}[h]
	\begin{subfigure}[t]{0.48\textwidth}
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
			\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
			\label{traverse_lock_a}	
		\end{subfigure}
		\\ \\ 
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
			\caption{The \lstinline{insert} method of the lock-coupling template algorithm written in C}
			\label{insert_lock}	
		\end{subfigure}
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		\label{traverse_lock_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms}
	\label{traverse_lock}
\end{figure}

@@

\than{I moved \texttt{insert} code of lock-coupling here}

Figure \ref{insert_lock} shows the implementation of \lstinline{insert} for the lock-coupling template. %\wm{Actually, is this \lstinline{insert} function part of the template? If so, should we discuss it in section 3?} 
It makes use of the \texttt{traverse} function %of Section \ref{lock-coupling-algo} 
to find the node at which to insert the key \lstinline{k}. If \texttt{traverse} returns 0, it has reached a node containing key \texttt{k}, and we only have to change the node's value to the target \texttt{value} (lines 5-6 in Figure \ref{insert_lock}). Otherwise, \texttt{traverse} has reached an empty leaf that can hold key \texttt{k}, so it calls the data-structure-specific \texttt{insertOp} function to allocate a new node with key \texttt{k} and value \texttt{v} (line 9 in Figure \ref{insert_lock}). \than{... that we will discuss in Section Verified DS.}

@@

\than{I moved \texttt{insert} from Section 4 to here. William, could you please write something to ensure a smooth transition between \texttt{traverse} and \texttt{insert}, or do you think it would be better to have subsections for \texttt{traverse} and \texttt{insert}? }

%\begin{figure}[h]
%	\begin{subfigure}[t]{0.48\textwidth}
%		\lstinputlisting[language=C, style=myStyle, escapechar=|]{lock_insert.c} 
%		\caption{The \lstinline{insert} method of the lock-coupling template algorithm}
%		\label{insert_lock}	
%	\end{subfigure}\qquad
%	\begin{subfigure}[t]{0.48\textwidth}
%		\lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
%		\caption{The \lstinline{insert} method of the give-up template algorithm}
%		\label{insert_giveup}
%	\end{subfigure} 
%	\caption{The \lstinline{insert} method of the lock-coupling and give-up template algorithms}
%	\label{insert_lock_giveup} 
%\end{figure}
%
%\than{Should we move the code to template section?}

@@

Figure \ref{proof_lock_insert} in Appendix \ref{sec:apd_proof} depicts the \texttt{insert} function within the lock-coupling template, annotated using separation logic.
The proof for \texttt{insert} begins with the precondition $\nodeboxrep \ \texttt{r}$. We set \texttt{pn->n} to point to the root node \texttt{*r} (line 3 in Figure \ref{proof_lock_insert}). After the lock for the root node is acquired (line 5 in Figure \ref{proof_lock_insert}), the thread acquires the contents of the root node, $\mathsf{R}(\texttt{r})$. We can then satisfy the precondition of the lock-coupling \lstinline{traverse} function from Section \ref{traverse_proof_lock} \wm{Maybe we should show the spec again here}, allowing us to call \lstinline{traverse} and obtain its postcondition.
\begin{mathpar}
	% spec of inRange
	{\color{blue}
		\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
		\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
		\right\rangle
	}
	\\ 
	\texttt{traverse(pn, k)} 
	\\
	{\color{blue}
		\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
		\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
			\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
		\end{array}
		\ \Bigg| \ \treerep\ m \
		\right\rangle
	}
\end{mathpar}
In the $\mathsf{then}$ branch, when \texttt{traverse} finds a node with key \lstinline{k}, %resulting in $\texttt{res} = \texttt{false}$, 
we know that $\texttt{pn->n'->t} \neq \texttt{NULL}$, i.e., the key to be inserted already exists in the tree. In this scenario, we update the node's value while maintaining the rest of the tree structure (line 9 in Figure \ref{proof_lock_insert}). Otherwise, when \texttt{traverse} reaches a leaf and produces $\mathit{res} = \texttt{true}$, we know that $\texttt{pn->n'->t} = \texttt{NULL}$ (at line 14-15 in Figure \ref{proof_lock_insert}). We can then call the \texttt{insertOp} specification, adding the new key-value pair to the tree. In either case, before we release the node's lock, we must show that we have altered the tree from its current abstract state $m$ to $m[\texttt{k} \mapsto \texttt{v}]$, thereby satisfying the public postcondition of \lstinline{insert}. We do this by demonstrating that the in-memory tree update corresponds to setting \lstinline{k} to \lstinline{v} in the abstract key-value map of the tree. 
We can then release the lock (line 21 in Figure \ref{proof_lock_insert}) return the node's contents to the shared state. %Subsequently, we demonstrate the postcondition of the \texttt{insert} function to complete the proof.
%\wm{Do we ever define $\treerep$?}

@@

%\begin{figure}[!ht]
%	\begin{subfigure}[t]{0.45\textwidth}
%		\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
%		\caption{The \lstinline{traverse} method of the give-up template algorithm written in an ML-like language}
%		\label{traverse_giveup_a}	
%	\end{subfigure}\qquad
%	\begin{subfigure}[t]{0.48\textwidth}
%		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
%		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
%		\label{traverse_giveup_b}
%	\end{subfigure}
%	\caption{The \lstinline{traverse} method of the give-up template algorithms}
%	\label{traverse_giveup}
%\end{figure}

\begin{figure}[h]
	\begin{subfigure}[t]{0.48\textwidth}
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
			\caption{The \lstinline{traverse} method of the give-up template algorithm written in an ML-like language}
			\label{traverse_giveup_a}	
		\end{subfigure}
		\\ \\ 
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
			\caption{The \lstinline{insert} method of the give-up template algorithm written in C}
			\label{insert_giveup}	
		\end{subfigure}
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
		\label{traverse_giveup_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} and \lstinline{insert} methods of the give-up template algorithms}
	\label{traverse_giveup}
\end{figure}

@@

\than{I moved the code of \texttt{insert} in give-up template here}

The give-up version of the \texttt{insert} method (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version. The one difference is that it does not require acquiring a lock before calling its \texttt{traverse} function.

@@

\than{I moved \texttt{insert} from Section 4 to here. William, could you please write something to ensure a smooth transition between \texttt{traverse} and \texttt{insert}, or do you think it would be better to have subsections for \texttt{traverse} and \texttt{insert}? }

 The proof of the \lstinline{insert} function is similar to that of the previous function: the only difference is that the give-up version of \lstinline{insert} does not acquire a lock before calling \lstinline{traverse}, reflecting the difference in the locking patterns and the preconditions of \lstinline{traverse} between the two templates. In Appendix \ref{sec:apd_proof}, Figure \ref{proof_giveup_insert} presents an annotation that uses separation logic for the \texttt{insert} function within the give-up template.

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% inRange for give-up
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{inRange(pn, k)} 
		\\
		{\color{blue}
			\left\{\mathit{res.} \
			\begin{array}{l}
				\texttt{pn->n}\mapsto\texttt{t} \ \ast \ \texttt{pn->n}\mapsto\texttt{min} \ \ast \  \texttt{pn->n}\mapsto\texttt{max} \  \ast  \ 
				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{min} < \texttt{k} < \texttt{max})\ \mathsf{else}\ (\texttt{k} \leq \texttt{min} \lor \texttt{k} \geq \texttt{max}))
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specifications of \texttt{inRange} for the give-up template}
	\label{fig:inRange_giveup}
\end{figure}
The give-up BST uses the same definition and specification of \texttt{findNext}, while introducing an extra helper function called \texttt{inRange}. %\wm{If \lstinline{inRange} is part of the template, I think we should discuss it in the previous section}. 
%The \texttt{insertOp} function is slightly different from that of the lock-coupling template, as each \lstinline{node_t} struct has two additional fields, \texttt{min} and \texttt{max}, to implement \lstinline{inRange}, as described in Section \ref{give-up-algo}.
The \texttt{inRange} function (Figure \ref{fig:inRange_giveup}) simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.

\subsection{Lock-coupling BST}
\label{BST_lock_insert}
\than{What do we need to say here? Which option should we choose here? (1) - Maintain the subsection and fill in \texttt{insertOp}, demonstrating how we prove \texttt{insert}, or (2) - Remove this subsection altogether.}
\subsection{Give-up BST}
\label{BST_giveup_insert}
\than{What do we need to say here? Which option should we choose here? (1) - Maintain the subsection and fill in \texttt{insertOp} and \texttt{inRange}, demonstrating how we prove \texttt{insert}, or (2) - Remove this subsection altogether.}

@@

Even though the proof for the \texttt{insert} function within the lock-coupling templates was line-of-code (LoC) intensive, future enhancements could significantly streamline this process. Otherwise, the construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, necessitated a considerable amount of time and effort.
However, we expect that functions such as \texttt{traverse}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.  The methods \texttt{findNext} and \texttt{insertOp} may also be adapted to other data structures, such as linked lists, with some additional effort. 
Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext}, \texttt{inRange}, and \texttt{insertOp} to restructure the \texttt{node} struct. While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future. The full development of our mechanization effort is available online
\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.
